Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2015-02-17 parts/intro.tex
fixes & typos

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Prevalent formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting~\cite{PMR10-TCSB},
consisting in a restriction of these frameworks where the evolution of a component is determined by the state of at most one other component that does not evolve.
In a sense, these kind of actions are of the form $X + Y \rightarrow X + Y'$ where $X$ behaves like a catalyst molecule that “hits” another molecule $Y$ and changes it into $Y'$, without being changed itself.
Assuming catalysts are always available, this can represent any biochemical system made of monomolecular reactions, and can also represent catalytic networks such as metabolic networks.
%, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
Our motivation behind this framework was to design a model and analysis techniques adapted to biological modelling.
These analyses avoid to build the whole state space, which allows to tackle very large systems (that would have led to a huge number of states, hopelessly too huge to be analysed).
They are based on the fact that most biological models have few levels of expression per component:
in Boolean networks~\cite{kauffman69,Thomas73} there are only two levels per component, and in their multivalued equivalent~\cite{deJong02}, components rarely have more than four levels.

@@

The approach used in this paper, 
consists in considering a broader class of models,
that we call Asynchronous Automata Network,
and that allows to naturally model these cooperations
by defining several requisites for a transition.
Moreover, such automata networks are still compatible with the notion of priority,
that can also be used to model different reaction rates in the model.
Asynchronous Automata Networks
(and, a fortiori, their restriction, the Process Hitting framework)
can be considered as a subset of Communicating Finite State Machines
or safe Petri Nets.
%or Synchronous Automata Networks,
%is inspired from the $\pi$-calculus,
Our work is thus related to such semantic ramifications
of extending traditional process algebras with the concept of priority
that allows for some transitions to be given precedence over others.
We focus here on static priorities that allow to model
time constraints such as reaction rates or delays between regulations,
but can also model preemptions between evolutions branches.

@@

\subsection*{Contributions}
In this paper, we develop a method
that allows to efficiently compute under-approximations of the dynamics
of Asynchronous Automata Networks (AANs),
generalising a prior work carried out on Process Hitting~\cite{PMR12-MSCS}.
Rather than using brute force or symbolic model checking techniques,
our method focuses on static analysis by abstract interpretation.
Our work aims at checking reachability properties that,
for a given initial state $s$ and a given local state $x$,
have the form:
“Starting from state $x$, is it possible to reach
a state in which $x$ is active?”,
thus answering either “True”
(in which case the reachability is formally proven)
or “Inconclusive”
(which however does not stand for “False”, that can be proven by other analysis tools,
such as the over-approximation proposed in~\cite{PMR12-MSCS}).
Moreover, the successive or simultaneous reachability of several local states
can also be checked,
and in the case of a “True” response,
an execution path satisfying the property can be produced.
Extensions of AANs where transitions are split in priority classes are addressed with an encoding in
AANs without classes of priorities.
Our work thus allows to efficiently analyse the dynamics
of regulation networks,
especially the widespread Logical Networks \cite{Thomas95,deJong02},
which encompass variables with a limited number of discrete values
alongside with evolution functions or focal parameters.
To show the scalability of our method, we apply it to a
large-scale biological models with around 100 components.

\pref{sec:ph} presents the Asynchronous Automata Networks (AANs)
without any notion of classes of priorities.
In \pref{sec:sa},
we develop our under-approximation method allowing to efficiently
compute reachability analysis;
we also show how to extract a valid execution path if the response is positive,
and propose two refinements in case one needs to check
successive or simultaneous reachability properties.
The framework of AANs with classes of priorities is given in \pref{sec:flattening},
alongside with an encoding into AANs without classes of priorities
(or, equivalently, with one class of priority).
Finally, \pref{sec:example} provides a detailed example and a large-scale example
of the application of our method,
and \pref{sec:ccl} gives a conclusion and a discussion about our work.

@@

\paragraph*{Sets}
If $A$ is a finite set,
$|A|$ is the cardinality of $A$
and $\powerset(A)$ is the power set of $A$.
$\sN$ is the set of natural numbers,
$\sN^* = \sN \setminus \{ 0 \}$ is the set of positive natural numbers,
$\sNN = \sN \setminus \{ 0, 1 \}$ is the set of natural numbers strictly greater than 1,
and $\segm{x}{y} = \{ x, x+1, \dots, y-1, y \}$ is the set of natural numbers from $x$ to $y$ included.
The Cartesian product of sets is denoted by $\times$;
if $z$ is a tuple of $n$ components, $\toset{z}$ denotes the corresponding set:
$\toset{z} = \{z_1, \cdots, z_n\}$.

\paragraph*{Sequences}
We denote by $\emptyseq$ the empty sequence.
If $n \in \sN$ and
$x = (x_i)_{i \in \segm{1}{n}}$ is a sequence of elements indexed by $i \in \segm{1}{n}$,
%we denote %$|x| = (b-a)+1$ the size of this sequence and
then $\indexes{x} = \segm{1}{n}$ is the set of indexes of this sequence.
Furthermore, if $a, b \in \indexes{x}$ with $a \leq b$,
then $x_{a..b} = (x_i)_{i \in \segm{a}{b}}$ is a subsequence of $x$.
%from element $n$ to element $m$ inclusive.
Finally, if $x$ is a sequence, $\toset{x}$ also denotes the corresponding set:
$\toset{x} = \{x_1, \cdots, x_{\card{x}}\}$.

\paragraph*{Digraph}
If $(V,E)$ is a directed graph whose $V$ are the nodes and $E \subseteq V\times V$ the edges,
the children of a node $n$ are given by
$\childs: V \to \powerset(V)$, with
$\childs(n) \DEF \{ m\in V\mid (n,m)\in E\}$;
its parents are given by
$\parents: V \to \powerset(V)$ with
$\parents(n) \DEF \{ m\in N\mid (m,n)\in E\}$;
and its successors are given by
$\conn_{(V,E)}(n)$ which is the least fixed point containing $n$
of the function
$\f{f}\conn:\powerset(V)\to \powerset(V)$
with $\f{f}\conn(W) \DEF \bigcup_{m\in W} \childs(m)$.

>>>>

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Prevalent formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting~\cite{PMR10-TCSB},
consisting in a restriction of these frameworks where the evolution of a component is determined by the state of at most one other component that does not evolve.
This is modelled by actions are of the form $X + Y \rightarrow X + Y'$ where $X$ behaves like a catalyst molecule that “hits” another molecule $Y$ and changes it into $Y'$, without being changed itself.
Assuming catalysts are always available, this can represent any biochemical system made of monomolecular reactions, and can also represent catalytic networks such as metabolic networks.
%, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
Our motivation behind this framework was to design a model and analysis techniques adapted to biological modelling.
These analyses avoid to build the whole state space, which allows to tackle very large systems (that would have led to a huge number of states, hopelessly too huge to be analysed).
They are based on the fact that most biological models have few levels of expression per component:
in Boolean networks~\cite{kauffman69,Thomas73} there are only two levels per component, and in their multivalued equivalent~\cite{deJong02}, components rarely have more than four levels.

@@

The approach used in this paper
consists in considering a broader class of models,
that we call Asynchronous Automata Network,
and that allows to naturally model these cooperations
by defining several requisites for a transition.
Moreover, such automata networks are still compatible with the notion of priority,
that can also be used to model different reaction rates in the model.
Asynchronous Automata Networks
(and, a fortiori, their restriction, the Process Hitting framework)
can be considered as a subset of Communicating Finite State Machines
or safe Petri Nets.
%or Synchronous Automata Networks,
%is inspired from the $\pi$-calculus,
Our work is thus related to such semantic ramifications
of extending traditional process algebras with the concept of priority
that allows for some transitions to be given precedence over others.
We focus here on static priorities that allow to model
time constraints such as reaction rates or delays between regulations,
but can also model preemptions between evolution branches.

@@

\subsection*{Contributions}
In this paper, we develop a method
that allows to efficiently compute under-approximations of the dynamics
of Asynchronous Automata Networks (AANs),
generalising a prior work carried out on Process Hitting~\cite{PMR12-MSCS}.
Rather than using brute force or symbolic model checking techniques,
our method focuses on static analysis by abstract interpretation.
Our work aims at checking reachability properties that,
for a given initial state $s$ and a given local state $x$,
have the form:
“Starting from state $x$, is it possible to reach
a state in which $x$ is active?”,
thus answering either “True”
(in which case the reachability is formally proven)
or “Inconclusive”
(which however does not stand for “False”, that can be proven by other analysis tools,
such as the over-approximation proposed in~\cite{PMR12-MSCS}).
Moreover, the successive or simultaneous reachability of several local states
can also be checked,
and in the case of a “True” response,
an execution path satisfying the property can be produced.
Extensions of AANs where transitions are split in priority classes are addressed with an encoding in
AANs without classes of priorities.
Our work thus allows to efficiently analyse the dynamics
of regulation networks,
especially the widespread Logical Networks \cite{Thomas95,deJong02},
which encompass variables with a limited number of discrete values
alongside with evolution functions or focal parameters.
To show the scalability of our method, we apply it to two
large-scale biological models with around 100 components.

\pref{sec:ph} presents the Asynchronous Automata Networks (AANs)
without any notion of classes of priorities.
In \pref{sec:sa},
we develop our under-approximation method allowing to efficiently
compute reachability analyses;
we also show how to extract a valid execution path if the response is positive,
and propose two refinements in case one needs to check
successive or simultaneous reachability properties.
The framework of AANs with classes of priorities is given in \pref{sec:flattening},
alongside with an encoding into AANs without classes of priorities
(or, equivalently, with one class of priority).
Finally, \pref{sec:example} provides a detailed example and two large-scale examples
of the application of our method,
and \pref{sec:ccl} gives a conclusion and a discussion about our work.

@@

\paragraph*{Sets}
If $A$ is a finite set,
$\card{A}$ is the cardinality of $A$
and $\powerset(A)$ is the power set of $A$.
$\sN$ is the set of natural numbers,
$\sN^* = \sN \setminus \{ 0 \}$ is the set of positive natural numbers
and $\segm{x}{y} = \{ x, x+1, \dots, y-1, y \}$ is the set of natural numbers from $x$ to $y$ included.
The Cartesian product of sets is denoted by $\times$ and
if $z$ is a tuple of $n$ components, $\toset{z}$ denotes the corresponding set:
$\toset{z} = \{z_1, \cdots, z_n\}$.

\paragraph*{Sequences}
We denote by $\emptyseq$ the empty sequence.
If $n \in \sN$ and
$x = (x_i)_{i \in \segm{1}{n}}$ is a sequence of elements indexed by $i \in \segm{1}{n}$,
then $\card{x} = n$ is the size of $x$
%we denote %$|x| = (b-a)+1$ the size of this sequence and
and $\indexes{x} = \segm{1}{n}$ is the set of indexes of this sequence.
Furthermore, if $a, b \in \indexes{x}$ with $a \leq b$,
then $x_{a..b} = (x_i)_{i \in \segm{a}{b}}$ is the subsequence of $x$
between indexes $a$ and $b$ included.
%from element $n$ to element $m$ inclusive.
Finally, if $x$ is a sequence, $\toset{x}$ also denotes the corresponding set:
$\toset{x} = \{x_1, \cdots, x_{\card{x}}\}$.

\paragraph*{Digraphs}
If $(V,E)$ is a directed graph whose set of nodes is $V$ and whose set of edges is $E \subseteq V\times V$,
the children of a node $n$ are given by
$\childs: V \to \powerset(V)$, with
$\childs(n) = \{ m\in V\mid (n,m)\in E\}$;
its parents are given by
$\parents: V \to \powerset(V)$ with
$\parents(n) = \{ m\in N\mid (m,n)\in E\}$;
and its successors are given by
$\conn_{(V,E)}(n)$ which is the least fixed point containing $n$
of the function
$\f{f}\conn:\powerset(V)\to \powerset(V)$
with $\f{f}\conn(W) = \bigcup_{m\in W} \childs(m)$.

### 2015-02-17 parts/models-ph-php.tex
fixes & typos

In this section, we define the notion of AANs with classes of priorities,
and give a translation from any of these AANs into
an AAN without priorities, as defined in \pref{sec:ph}.

The idea behind AANs with classes of priorities (\pref{def:php})
is to split the set of actions into several subsets assigned to priorities,
and adapting the behaviour of the model to make any action unplayable
until no other action of higher priority is playable (\pref{def:playp}).
Such model allows to model preemptions between sets of actions,
which can be helpful to abstract time or duration properties under certain conditions.

\begin{definition}[AAN with $k$ classes of priorities]
\label{def:php}
  If $k \in \sN^*$,
  an \emph{Asynchronous Automata Network with $k$ classes of priorities} (AAN$k$)
  is a triplet $\PH = (\PHs; \PHl; \PHa^{\angles{k}})$,
  where $\PH^{\angles{k}} = (\PHh^{(1)} \dots; \PHh^{(k)})$,
  and:
  \begin{itemize}
    \item $\PHs \DEF \{a, b, \dots, z\}$ is the finite set of \emph{automata};
    \item $\PHl \DEF \underset{a \in \PHs}{\times} \PHl_a$ is the finite set of
      (global) \emph{states},
      where $\PHl_a = \{a_0, \ldots, a_{l_a}\}$ is the finite set of \emph{local states}
      of automaton $a \in \PHs$, with $l_a \in \sN^*$,
      and so that:
      $\forall (a_i; b_j) \in \PHl_a \times \PHl_b, a \neq b \Rightarrow a_i \neq b_j$;
    \item $\forall n \in \llbracket 1; k \rrbracket,
      \PHa^{(n)} \DEF \{\PHfrappe{A}{b_j}{b_k} \mid
      b \in \PHs \wedge (b_j; b_k) \in \PHl_b \times \PHl_b \wedge
      b_j \neq b_k \wedge
      \forall a \in \PHs, \card{A \cap \PHl_a} \leq 1 \wedge
      A \cap \PHl_b = \emptyset \}$ is the finite set of \emph{actions of priority $n$}.
  \end{itemize}
  We also use the same notations as those defined in \pref{sec:ph}, when applicable.
  Furthermore,
  we denote by $\PHh \DEF \bigcup_{n \in \segm{1}{k}} \PHh^{(n)}$ the set of all actions
  and, for all $h \in \PHh$,
  by $\prio(h) \DEF \min\{ n \in \segm{1}{k} \mid h \in \PHh^{(n)} \}$
  the priority of action $h$.
\end{definition}

\begin{definition}[Semantics of an AAN$k$ ($\PHPtrans$)]
\label{def:playp}
  An action $h = \PHhit{A}{b_j}{b_k} \in \PHa^{(n)}$ of priority $n$
  is \emph{playable} in $s \in \PHl$
  if and only if $A \subseteq s$, $\PHget{s}{b} = b_j$ and
  $\forall m < n, \forall g \in \PHa^{(m)},
    \neg (\PHhitter(g) \subseteq s \wedge \PHtarget(g) \in s)$.
  In such a case, $(s \PHplay h)$ stands for the state resulting from the play
  of the action $h$ in $s$, which is defined by: 
    $(s \PHplay h)_b = b_k$
	and
	$\forall a\in\PHs, a\neq b, (s\PHplay h)_a = s_a$.
  Moreover, we denote: $s \PHPtrans (s \PHplay h)$.
\end{definition}

We consider in the following an AAN$k$ with $k \in \sNN$:
$\ov{\PH} = (\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}})$.
The aim of the rest this section is to propose a translation of $\oPH$
into an AAN with $1$ class of priority $\PH = (\PHs; \PHl; \PHa^{\angles{1}})$
called \emph{flattening}, which is bisimilar.
This AAN with 1 class of priority is equivalent to a regular AAN without priorities;
Therefore, such a translation is particularly useful to be able to study the dynamics of
any kind of AAN with priorities by using the static analysis developed in \pref{sec:sa}.

The translation given in the following relies on the notion of \emph{sub-state},
which is a set of local states containing at most one local state of each automata.
Thus, a sub-state can be considered as a partial state.
This notion if formally expressed in \pref{def:substate}.

\begin{definition}[Sub-state ($\PHsublize{\PHl}$)]
\label{def:substate}
  If $S \subseteq \PHs$ is a set of automata, a sub-state on $S$ is an element of:
  $\PHsubl[\PHl]_S \DEF \{ \toset{\rho} \subseteq \Proc \mid
    \rho \in \bigtimes{a \in S} \PHl_a \}$
  where the notation $\toset{\rho}$ represents
  the set of components of the Cartesian product $\rho$,
  as defined on page~\pageref{notations}.
  The set of all sub-states is denoted by:
  $\PHsubl[\PHl] \DEF \bigcup_{S \in \powerset(\PHs)} \PHsubl[\PHl]_S$.
  Furthermore, if $\mysigma \in \PHsubl[\PHl]$ and $s \in \PHl$, we note:
    \[\mysigma \subseteq s \EQDEF \forall a_i \in \Proc, a_i \in \mysigma \Rightarrow a_i \in s\]
\end{definition}

The translation of an AAN$k$ into an AAN
is based on the notion of \emph{playability property} defined in \pref{def:pp}
which is a Boolean formula where the atoms are local states of $\ov{\PH}$.

\begin{definition}[Playability property language ($\F$)]
  \label{def:pp}
  A \emph{playability property} is an element of the language $\F$ inductively defined by:
  \begin{itemize}
    \item $\top$ and $\bot$ belong to $\F$;
    \item if $a \in \ov{\PHs}$ and $a_i \in \ov{\PHl}_a$, then $a_i \in \F$ and we call it an \emph{atom};
    \item if $P \in \F$ and $Q \in \F$, then $\neg P \in \F$, $P \wedge Q \in \F$ and $P \vee Q \in \F$.
  \end{itemize}
  If $P \in \F$ is a playability property and $\mysigma \in \PHsubl$ is a sub-state of $\oPH$,
  we note $\Fsem{P}{\mysigma}$ the \emph{evaluation} of $P$ in $\mysigma$:
  \begin{itemize}
    \item if $P = a_i \in \ov{\PHl}_a$ is an atom, with $a \in \ov{\PHs}$, then $\Fsem{a_i}{\mysigma}$ is true iff $a_i \in \mysigma$;
    \item if $P$ is not an atom, then $\Fsem{P}{\mysigma}$ is true iff all its subformulas are inductively true in $\mysigma$
      with the classical semantics for the logic operators $\top$, $\bot$, $\neg$, $\wedge$ and $\vee$.
  \end{itemize}
\end{definition}

Because we only use classical logic operators, the formulas of Boolean logic on 
distributivity, associativity and commutativity can be used, together with De Morgan's laws on negation.
We also have the following property for the negation of an atom:
\[\forall a \in \ov{\PHs}, \forall a_i \in \ov{\PHl}_a, \forall \mysigma \in \PHsubl,
  \Fsem{\neg a_i}{\mysigma} \Leftrightarrow \Fsem{\bigvee_{\substack{a_j \in \ov{\PHl}_a\\a_j \neq a_i}} a_j}{\mysigma}\]
Indeed, if a local state is not active in a state, this means that another local state of the same automaton is active.
Moreover, we note that:
\[\forall a \in \ov{\PHs}, \forall a_i, a_j \in \ov{\PHl}_a,
  a_i \neq a_j \Longrightarrow a_i \wedge a_j \equiv \bot \]
because two different processes can never be active simultaneously.

In \pref{def:fop}, we define the the operator $\Fopsymbol$ which characterises the playability of an action
given the semantics of AAN$k$s (see \pref{def:playp}).
This operator simply states that the hitters of an action have to be active,
and no other action of higher priority have to be playable.
% The target of the considered action is not taken into account
% as it is not needed for the rest of the flattening.

\begin{definition}[Playability property operator ($\Fopsymbol : \PHh \rightarrow \F$)]\label{def:fop}
  For all $h = \PHfrappe{A}{b_j}{b_m} \in \ov{\PHh}$, we define:
  \[
    \Fop{h} \equiv
    b_j \wedge
    \left(\bigwedge_{a_i \in A} a_i\right)
    %\hitter{h}
    \wedge
      \left( \bigwedge_{\substack{g \in \ov{\PHh}^{(n)}\\1 < n < \prio(h)}}
      \neg \left( \target{g} \wedge \left(\bigwedge_{c_l \in \hitter{g}} c_l\right)\right) \right)
  \]
\end{definition}
%
By construction of this operator and given the dynamics of a AAN$k$,
an action $h$ is playable in a state $s \in \ov{\PHl}$ if and only if: $\Fsem{\Fop{h}}{s}$.

@@

\begin{lemma}
\label{lem:ppplaysubset}
  Let $h \in \ov{\PHh}$ and $s \in \ov{\PHl}$;
  $h$ is playable in $s$ if and only if:
  \[\target{h} \in s \wedge \exists  \mysigma \in \PHsubl,
    \mysigma \subseteq s \wedge \Fsem{\Fop{h}}{\mysigma} \enspace.\]
\end{lemma}
%
\begin{proof}
  ($\Rightarrow$)
    If $h$ is playable in $s$, then $\target{h} \in s$ and $\Fsem{\Fop{h}}{s}$.
    Thus, $\Fop{h} \not\equiv \bot$ and, by property of a DNF,
    at least one of the $\n$ conjunctions of $\Fop{h}$ is true in $s$.
    Suppose the $i$\textsuperscript{th} conjunction is true in $s$, with $i \in \segm{1}{\n}$;
    then we have: $\forall j \in \segm{1}{\m}, p_{i,j} \in s$.
    Let $\mysigma \in \PHsubl_{\PHdep{i}{h}}$
    with $\forall b \in \PHdep{i}{h}, \PHget{\mysigma}{b} = \PHget{s}{b}$.
    We immediately have: $\mysigma \subseteq s$,
    and, by construction of $\PHdep{i}{h}$, $\Fsem{\Fop{h}}{\mysigma}$.

@@

\begin{definition}[Flattening ($\PHflat$)]
  \label{def:flattening}
  If $k \in \sNN$ and $\oPH = (\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}})$ is an AAN$k$,
  we denote by
  $\PHflat(\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}}) = (\PHs; \PHl; \PHa)$
  the \emph{flattening} of $\oPH$, where:
  \begin{itemize}
    \item $\PHs = \ov{\PHs}$;

>>>>

In this section, we define the notion of AANs with classes of priorities,
and give a translation from these into
AANs without priorities, as defined in \pref{sec:ph}.

The idea behind AANs with classes of priorities (\pref{def:php})
is to split the set of actions into several subsets assigned to priorities,
and to adapt the behaviour of the model to make any action unplayable
until no other action of higher priority is playable (\pref{def:playp}).
Such model allows to model preemptions between sets of actions,
which can be helpful to abstract time or duration properties under certain conditions.

\begin{definition}[AAN with $k$ classes of priorities]
\label{def:php}
  If $k \in \sN^*$,
  an \emph{Asynchronous Automata Network with $k$ classes of priorities} (AAN$k$)
  is a triplet $\PH = (\PHs; \PHl; \PHa^{\angles{k}})$,
  where $\PH^{\angles{k}} = (\PHh^{(1)} \dots; \PHh^{(k)})$,
  and:
  \begin{itemize}
    \item $\PHs \DEF \{a, b, \dots, z\}$ is the finite set of \emph{automata};
    \item $\PHl \DEF \underset{a \in \PHs}{\times} \PHl_a$ is the finite set of
      (global) \emph{states},
      where $\PHl_a = \{a_0, \ldots, a_{l_a}\}$ is the finite set of \emph{local states}
      of automaton $a \in \PHs$, with $l_a \in \sN^*$,
      and so that:
      $\forall (a_i; b_j) \in \PHl_a \times \PHl_b, a \neq b \Rightarrow a_i \neq b_j$;
    \item $\forall n \in \llbracket 1; k \rrbracket,
      \PHa^{(n)} \DEF \{\PHfrappe{A}{b_j}{b_k} \mid
      b \in \PHs \wedge (b_j; b_k) \in \PHl_b \times \PHl_b \wedge
      b_j \neq b_k \wedge
      \forall a \in \PHs, \card{A \cap \PHl_a} \leq 1 \wedge
      A \cap \PHl_b = \emptyset \}$ is the finite set of \emph{actions of priority $n$}.
  \end{itemize}
\end{definition}

\paragraph{Notations}
We use the same notations as those defined in \pref{sec:ph}, when applicable.
Furthermore,
we denote by $\PHh = \bigcup_{n \in \segm{1}{k}} \PHh^{(n)}$ the set of all actions
and, for all $h \in \PHh$,
by $\prio(h) = \min\{ n \in \segm{1}{k} \mid h \in \PHh^{(n)} \}$
the priority of action $h$.

\begin{definition}[Semantics of an AAN$k$ ($\PHPtrans$)]
\label{def:playp}
  An action $h = \PHhit{A}{b_j}{b_k} \in \PHa^{(n)}$ of priority $n$
  is \emph{playable} in $s \in \PHl$
  if and only if $A \subseteq s$, $\PHget{s}{b} = b_j$ and
  $\forall m < n, \forall g \in \PHa^{(m)},
    \neg (\PHhitter(g) \subseteq s \wedge \PHtarget(g) \in s)$.
  In such a case, $(s \PHplay h)$ stands for the state resulting from the play
  of the action $h$ in $s$, which is defined by: 
    $\PHget{(s \PHplay h)}{b} = b_k$
  and
    $\forall a \in \PHs, a \neq b, \PHget{(s\PHplay h)}{a} = \PHget{s}{a}$.
  Moreover, we denote: $s \PHPtrans (s \PHplay h)$.
\end{definition}

We consider in the following an AAN$k$:
$\ov{\PH} = (\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}})$
with $k \in \sN$, $k > 1$.
The aim of the rest this section is to propose a translation of $\oPH$
into an AAN with $1$ class of priority $\PH = (\PHs; \PHl; \PHa^{\angles{1}})$
called \emph{flattening}, which is bisimilar.
As an AAN with 1 class of priority is equivalent to a regular AAN without priorities,
such a translation is particularly useful to be able to study the dynamics of
any kind of AAN with priorities by using the static analysis developed in \pref{sec:sa}.

The translation given in the following relies on the notion of \emph{sub-state}
(\pref{def:substate}),
which is a set of local states containing at most one local state of each automata.
Thus, a sub-state can be considered as a partial state.

\begin{definition}[Sub-state ($\PHsublize{\PHl}$)]
\label{def:substate}
  If $S \subseteq \PHs$ is a set of automata, a sub-state on $S$ is an element of:
  $\PHsubl[\PHl]_S \DEF \{ \toset{\rho} \subseteq \Proc \mid
    \rho \in \bigtimes{a \in S} \PHl_a \}$
  (where the notation $\toset{\rho}$ represents
  the set of components of the Cartesian product $\rho$,
  as defined on page~\pageref{notations}).
  The set of all sub-states is denoted by:
  $\PHsubl[\PHl] \DEF \bigcup_{S \in \powerset(\PHs)} \PHsubl[\PHl]_S$.
  Furthermore, we recall the notation from \pref{sec:ph},
  if $\mysigma \in \PHsubl[\PHl]$ and $s \in \PHl$:
    \[\mysigma \subseteq s \Leftrightarrow \forall a_i \in \mysigma, \PHget{s}{a} = a_i\]
%    \[\mysigma \subseteq s \EQDEF \forall a_i \in \Proc, a_i \in \mysigma \Rightarrow a_i \in s\]
\end{definition}

The translation of an AAN$k$ into an AAN
is based on the notion of \emph{playability property} (\pref{def:pp})
which is a Boolean formula where the atoms are local states of $\ov{\PH}$.

\begin{definition}[Playability property language ($\F$)]
  \label{def:pp}
  A \emph{playability property} is an element of the language $\F$ inductively defined by:
  \begin{itemize}
    \item $\top$ and $\bot$ belong to $\F$;
    \item if $a \in \ov{\PHs}$ and $a_i \in \ov{\PHl}_a$, then $a_i \in \F$ and we call it an \emph{atom};
    \item if $P \in \F$ and $Q \in \F$, then $\neg P \in \F$, $P \wedge Q \in \F$ and $P \vee Q \in \F$.
  \end{itemize}
  If $P \in \F$ is a playability property, we denote by: $\PHsortf{P}$
  the set of sorts whom at least one local state is mentioned in $P$.
  Then, if $\mysigma \in \PHsubl$
  with $\PHsortf{P} \subseteq \PHsortf{\mysigma}$,
  we note $\Fsem{P}{\mysigma}$ the \emph{evaluation} of $P$ in $\mysigma$:
  \begin{itemize}
    \item if $P = a_i \in \ov{\PHl}_a$ is an atom, with $a \in \ov{\PHs}$, then $\Fsem{a_i}{\mysigma}$ is true iff $a_i \in \mysigma$;
    \item if $P$ is not an atom, then $\Fsem{P}{\mysigma}$ is evaluated in $\mysigma$
      with the classical semantics for the logic operators $\top$, $\bot$, $\neg$, $\wedge$ and $\vee$.
  \end{itemize}
\end{definition}

Because we only use classical logic operators, the formulas of Boolean logic on 
distributivity, associativity and commutativity can be used, together with De Morgan's laws on negation.
We also have the following property for the negation of an atom:
% \[\forall a \in \ov{\PHs}, \forall a_i \in \ov{\PHl}_a, \forall \mysigma \in \PHsubl,
%   a \in \PHsortf{\mysigma},
%   \Fsem{\neg a_i}{\mysigma} \Leftrightarrow \Fsem{\bigvee_{\substack{a_j \in \ov{\PHl}_a\\a_j \neq a_i}} a_j}{\mysigma}\]
\[\forall a \in \ov{\PHs}, \forall a_i \in \ov{\PHl}_a,
  \neg a_i \Longleftrightarrow \bigvee_{\substack{a_j \in \ov{\PHl}_a\\a_j \neq a_i}} a_j\]
Indeed, if a local state is not active in a state, this means that another local state of the same automaton is active.
Moreover, we note that:
\[\forall a \in \ov{\PHs}, \forall a_i, a_j \in \ov{\PHl}_a,
  a_i \neq a_j \Longrightarrow a_i \wedge a_j \equiv \bot \]
because two different processes can never be active simultaneously.

In \pref{def:fop}, we define the the operator $\Fopsymbol$ which characterises the playability of an action
given the semantics of AAN$k$s (see \pref{def:playp}).
This operator simply states that the hitters of an action have to be active,
and every other action of higher priority has to be not playable.
% The target of the considered action is not taken into account
% as it is not needed for the rest of the flattening.

\begin{definition}[Playability property operator ($\Fopsymbol : \PHh \rightarrow \F$)]\label{def:fop}
  For all $h = \PHfrappe{A}{b_j}{b_m} \in \ov{\PHh}$, we define:
  \[
    \Fop{h} \equiv
    b_j \wedge
    \left(\bigwedge_{a_i \in A} a_i\right)
    %\hitter{h}
    \wedge
      \left( \bigwedge_{\substack{g \in \ov{\PHh}^{(n)}\\1 \leq n < \prio(h)}}
      \neg \left( \target{g} \wedge \left(\bigwedge_{c_l \in \hitter{g}} c_l\right)\right) \right)
  \]
\end{definition}
%
By construction of this operator and given the dynamics of a AAN$k$,
an action $h$ is playable in a state $s \in \ov{\PHl}$ if and only if: $\Fsem{\Fop{h}}{s}$.

@@

\begin{lemma}
\label{lem:ppplaysubset}
  Let $h \in \ov{\PHh}$ and $s \in \ov{\PHl}$;
  $h$ is playable in $s$ if and only if:
  \[\target{h} \in s \wedge \exists \n \in \sN, \exists \mysigma \in \PHsubl_{\PHdep{i}{h}},
    \mysigma \subseteq s \wedge \Fsem{\Fop{h}}{\mysigma} \enspace.\]
\end{lemma}
%
\begin{proof}
  ($\Rightarrow$)
    If $h$ is playable in $s$, then $\target{h} \in s$ and $\Fsem{\Fop{h}}{s}$.
    Thus, $\Fop{h} \not\equiv \bot$ and, by property of a DNF,
    at least one of the $\n$ conjunctions of $\Fop{h}$ is true in $s$.
    Suppose the $i$\textsuperscript{th} conjunction is true in $s$, with $i \in \segm{1}{\n}$;
    then we have: $\forall j \in \segm{1}{\m}, p_{i,j} \in s$.
    Let $\mysigma \in \PHsubl_{\PHdep{i}{h}}$
    with $\forall b \in \PHdep{i}{h}, \PHget{\mysigma}{b} = \PHget{s}{b}$.
    We immediately have: $\mysigma \subseteq s$,
    and, by construction of $\PHdep{i}{h}$, $\Fsem{\Fop{h}}{\mysigma}$.

@@

\begin{definition}[Flattening ($\PHflat$)]
  \label{def:flattening}
  If $k \in \sNN$, $k > 1$ and $\oPH = (\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}})$ is an AAN$k$,
  we denote by
  $\PHflat(\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}}) = (\PHs; \PHl; \PHa)$
  the \emph{flattening} of $\oPH$, where:
  \begin{itemize}
    \item $\PHs = \ov{\PHs}$;
